Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-05-20 paper/paper.tex
Updates from Overleaf

Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small scale spatial variability. The broad bands of multi-spectral imagers coupled with the irregular shape of lakes and rivers results in pixels with highly mixed signals that are easily dominated by reflectance from shore and near-shore vegetation sources \cite{koponen2002lake, ritchie2003remote}. Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAV) such as drones has emerged as a powerful approach to simultaneously address the spectral, spatial, and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}. UAVs are significantly less expensive to deploy than their remote sensing counterparts, and low altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}. However, the significant increase in data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.

Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters from UAV-captured hyperspectral images (HSI). On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}. These band ratios and polynomial combinations of bands have been used to successfully invert optically-active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}. Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a plethora of parameters such as colored dissolved organic matter, chlorophyll a, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}. The calibration and evaluation of these data-driven models demand a significant volume of coincident, in situ data. This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}. Nevertheless, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation. The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{FIXME}

\funding{FIXME}

@@

\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}

>>>>

Inland water bodies present a unique challenge to characterization by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability. The broad bands of multispectral imagers coupled with the irregular shape of lakes and rivers result in pixels with highly mixed signals that are easily dominated by reflectance from shore and nearshore vegetation sources \cite{koponen2002lake, ritchie2003remote}. Recently, the combination of hyperspectral imaging with unmanned aerial vehicles (UAVs), such as drones, has emerged as a powerful approach to simultaneously address the spectral, spatial and temporal limitations of traditional high-altitude and satellite-based collection \cite{adao2017hyperspectral,arroyo2019implementation}. UAVs are significantly less expensive to deploy than their satellite or aircraft based remote sensing counterparts, and low-altitude flights enable centimeter-scale sampling while limiting the need for complicated atmospheric corrections \cite{banerjee2020uav}. However, the significant increase in the data volume generated by these systems presents a new challenge, namely, how to efficiently extract water quality parameters of interest from intricate pixel spectra.

Significant research efforts have focused on the development of techniques and algorithms to retrieve water quality parameters, e.g. from UAV-captured hyperspectral images (HSI). On-board compute installed alongside hyperspectral imagers can enable the rapid evaluation of spectral indices from HSI band ratios \cite{horstrand2019uav}. These band ratios and polynomial combinations of bands have been used to successfully invert optically active water quality parameters such as turbidity directly from UAV acquired imagery \cite{vogt2016near, zhang2022selection}. Supervised machine learning techniques such as tree-based models, support vector machines, and neural networks have also been used to estimate a wide range of parameters such as colored dissolved organic matter, chlorophyll A, blue-green algae, and suspended sediment concentrations \cite{keller2018hyperspectral, lu2021retrieval}. The calibration and evaluation of these data-driven models require a significant volume of coincident in situ data. This can be addressed by coordinating UAV flights with reference data collection using autonomous robotic boats \cite{robot-team-1, robot-team-2}. However, this approach relies on prior knowledge of expected sources in order to select appropriate reference instruments for model validation. The presence of unanticipated contaminants cannot be directly identified in this sensing paradigm.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\authorcontributions{Methodology, J.W. and D.J.L; conceptualization, D.J.L.; software, J.W.; field deployment and preparation, J.W., A.A., L.O.H.W., S.T., B.F., P.M.H.D., M.I., M.L., D.S., and D.J.L.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; data curation, J.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.W.; writing---review and editing, J.W. and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L. All authors have read and agreed to the published version of the manuscript.
}

\funding{This research was funded by the following grants: the Texas National Security Network Excellence Fund award for Environmental Sensing Security Sentinels; the SOFWERX award for Machine Learning for Robotic Teams %Please ensure meaning has been retained. 
	and NSF Award OAC-2115094; support from the University of Texas at Dallas Office of Sponsored Programs, Dean of Natural Sciences and Mathematics, and Chair of the Physics Department is gratefully acknowledged; TRECIS CC* Cyberteam (NSF \#2019135); NSF OAC-2115094 Award; and EPA P3 grant number 84057001-0.
}

@@

\acknowledgments{Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.  Christopher Simmons is gratefully acknowledged for his computational support. We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite. Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage. Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team. The authors acknowledge the OIT-Cyberinfrastructure Research Computing group at the University of Texas at Dallas and the TRECIS CC* Cyberteam (NSF \#2019135) for providing HPC resources that contributed to this research; the authors also acknowledge their receipt of the %Please ensure meaning has been retained. 
	NSF OAC-2115094 Award and EPA P3 grant number 84057001-0.}

\conflictsofinterest{The authors declare no conflicts of interest.}
